---
freeze: true
---

# Binary data formats {#sec-binary-data}

```{r setup}
#| results: "asis"
#| echo: false
source("_common.R")
status("drafting")
```

## Introduction

### Prerequisites

This chapter focuses on large data sets, and binary file formats that are well suited to working with them.
We'll use a data set of item checkouts from Seattle public libraries, available online at [data.seattle.gov/Community/Checkouts-by-Title/tmmm-ytt6](https://data.seattle.gov/Community/Checkouts-by-Title/tmmm-ytt6).
The data is distributed as a single 9GB csv file, so it may take some time to download.
For the purposes of this chapter we'll assume that you have a local copy of the file at `data/seattle-library-checkout.csv`.

```{r load-packages}
#| message: false
#| warning: false
library(tidyverse)
library(arrow)
library(duckdb)
library(dbplyr)
```

## The arrow package

In this chapter we'll rely on Apache Arrow ([arrow.apache.org](https://arrow.apache.org/)), a multi-language toolbox designed for efficient analysis and transport of large data sets.
The arrow package provides a standard way to use Apache Arrow in R, and allows you to analyze larger-than-memory datasets using familiar dplyr syntax.
As an additional benefit, arrow is extremely fast: you'll see some examples later in the chapter.

### Opening a dataset

Let's start by taking a look at the Seattle library checkout data.
At 9GB in size, the csv file is large enough that we probably don't want to load the whole thing into memory: a good rule of thumb is that you usually want at least twice as much memory as the size of the data.
Instead of using `read_csv()` we'll use the `open_dataset()` function from the arrow package:

```{r open-dataset}
seattle_lib <- open_dataset(
  sources = "data/seattle-library-checkouts.csv", 
  format = "csv"
)
```

When this code is run, `open_dataset()` scans the file to work out its structure and creates the `seattle_lib` object that stores metadata.
The data remain on-disk, and are not loaded into memory.
Nevertheless, we are able to call dplyr functions to interact with the data:

```{r glimpse-data}
#| cache: true
glimpse(seattle_lib)
```

The first line in the output tell you that `seattle_lib` is stored locally on-disk as a single csv file.
The second line indicates that the data are stored as a tabular data set with 41 million rows and 12 columns.

### Using dplyr with arrow

Writing dplyr code for arrow data is conceptually similar to dbplyr, described in @sec-import-databases.
The dplyr code that you write in R is internally transformed to a query that the Apache Arrow C++ library understands, which is then executed when you call `collect()`.
Here's an example that counts the total number of books checked out in each year:

```{r books-by-year}
#| cache: true
seattle_lib |> 
  filter(MaterialType == "BOOK") |>
  group_by(CheckoutYear) |>
  summarise(TotalCheckouts = sum(Checkouts)) |>
  arrange(desc(CheckoutYear)) |>
  collect()
```

Like dbplyr, arrow only understands some R expressions, so you may not be able to write exactly the same code you usually would.
However, the list of operations and functions supported is fairly extensive and growing.
A complete list is provided in the arrow package documentation at [arrow.apache.org/docs/r/reference/acero.html](https://arrow.apache.org/docs/r/reference/acero.html).

## The parquet file format

When working with small data sets, csv files can be convenient: they're human readable plain text files, and most analysis tools can read data from a csv.
However, there are some drawbacks:

-   Ambiguity.
    A csv file does not uniquely specify the data structure.
    For example, a csv reader has to guess whether `"2022-10-08"` should be parsed as a string or a date.
    If it is a date, it has to guess whether the date refers to October 8th or August 10th.
    Here is a slightly contrived example.
    Here is a small tibble that contains some integers and some strings that look suspiciously like dates:

    ```{r csv-ambiguity-example}
    original <- tibble(
      integers = 1L:3L,
      strings = c("2022-01-01", "1980-12-01", "1900-12-12")
    )
    original
    ```

    Simple though it is, this data set poses problems for a csv file.
    The csv version of this data does not explicitly specify that the first column contains integers and the second column contains strings.
    This produces ambiguity because the csv reader guesses that the data types are double and character respectively:

    ```{r csv-ambiguity-failure}
    csv_file <- tempfile(fileext = ".csv")
    write_csv(original, csv_file)
    read_csv(csv_file, show_col_types = FALSE)
    ```

    The `read_csv()` and `write_csv()` functions are designed to avoid this issue to an extent, but it is impossible to avoid it completely because the csv format is inherently ambiguous.

-   Inaccuracy.
    Plain text files store binary numeric data inaccurately due to precision limits.
    We can verify this by writing 1000 random numbers to a csv file and then reloading the data from the file

    ```{r csv-precision-failure}
    #| message: false
    csv_file <- tempfile(fileext = ".csv")
    original <- tibble(old_x = rnorm(1000))
    write_csv(original, csv_file)
    reloaded <- read_csv(csv_file) |>
      rename(new_x = old_x)
    ```

    Now let's compare the two, counting the number of times the original value is strictly identical to the reloaded one:

    ```{r csv-precision-failure-2}
    bind_cols(original, reloaded) |> 
      mutate(same_value = old_x == new_x) |>
      count(same_value)
    ```

    A substantial proportion of the reloaded values differ from the original ones.
    These differences are very small in magnitude, and occur because the csv represents numeric values with limited precision.
    Even so this can still be a source of problems in your analysis code.

-   Poor performance.
    Compared to other formats, csv files can be slow to read and write because the on-disk format is very different to how the data needs to be represented in-memory.
    We'll see some examples of this later in the chapter.

The Apache Parquet file format ([parquet.apache.org](https://parquet.apache.org/)) is designed to address these limitations.
Like a csv file, a parquet file stores a single rectangular data set.
Features of the parquet format that make it appealing for data scientists include:

-   Parquet files store data in a binary format (not plain text), avoiding inaccuracies in encoding and making it faster to read and write than a csv file.
-   Parquet files store metadata that uniquely identifies the data types stored in each column, which avoids the ambiguity of csv files.
-   Parquet files are compressed by default, and are typically much smaller than the corresponding csv files would be.
-   Parquet files internally break the data into distinct chunks, each with associated metadata. This makes it possible to write intelligent file readers that can jump straight to a specific part of the data when the full data set is not needed.

All these factors together make the parquet format appealing when working with larger data sets.

The arrow package supplies `read_parquet()` and `write_parquet()` functions that can read and write data in the parquet format.
We'll use these functions to explore the advantages to parquet files.

### Parquet files are unambiguous

Recall this example from earlier:

```{r ambiguous-data-frame}
original <- tibble(
  integers = 1L:3L,
  strings = c("2022-01-01", "1980-12-01", "1900-12-12")
)
original
```

As we saw, writing this data to a csv file and reading it back in leads to a mistake in identifying the data types for each column.
The parquet format avoids this ambiguity.
When we write the data set to a parquet file using `write_parquet()` and then read it back in with `read_parquet()`, the data types are preserved:

```{r read-parquet-unambigiously}
parquet_file <- tempfile(fileext = ".parquet")
write_parquet(original, parquet_file)
read_parquet(parquet_file)
```

### Parquet files are binary

Earlier in this section we saw an example in which csv files store numeric data inaccurately because they use plain text rather than binary format.
Let's repeat the exercise using the binary parquet format.
We first write some random numbers to a parquet file:

```{r parquet-precision-write}
#| message: false
parquet_file <- tempfile(fileext = ".parquet")
original <- tibble(old_x = rnorm(1000))
write_parquet(original, parquet_file)
```

Next we load the file and give the reloaded variable a new name:

```{r parquet-precision-read}
reloaded <- read_parquet(parquet_file) |>
  rename(new_x = old_x)
```

Finally we bind the old and new data sets together and compare them:

```{r parquet-precision-success}
bind_cols(original, reloaded) |> 
  mutate(same_value = old_x == new_x) |>
  count(same_value)
```

As you can see, numeric values stored in parquet files are stored faithfully.
Loading the data from file produces the same values in the original data set.

### Parquet files are smaller

We can use the Seattle library data set to highlight the advantage in file size.
The `write_parquet()` function can take an Arrow Dataset object like `seattle_lib` as the input, so we can write the file like this:

```{r write-parquet}
#| eval: !expr "!file.exists('data/seattle-library-checkouts.parquet')"
write_parquet(seattle_lib, "data/seattle-library-checkouts.parquet")
```

Now let's compare the two files in size:

```{r compare-file-sizes}
file.size("data/seattle-library-checkouts.csv")
file.size("data/seattle-library-checkouts.parquet")
```

The parquet file is about half the size of the csv file.

### Parquet files are faster

The format of your data file can have a substantial impact on how long a data analysis pipeline takes to execute.
To see this, let's time how long it takes to count the number of books checked out in each month of 2021 when the data are stored in csv format:

```{r dataset-performance-csv}
#| cache: true
system.time({
  open_dataset(
    sources = "data/seattle-library-checkouts.csv", 
    format = "csv"
  ) |>
    filter(CheckoutYear == 2021, MaterialType == "BOOK") |>
    group_by(CheckoutMonth) |>
    summarise(TotalCheckouts = sum(Checkouts)) |>
    arrange(desc(CheckoutMonth)) |>
    collect()
})
```

Now let's repeat the exercise using the parquet file:

```{r dataset-performance-parquet}
#| cache: true
system.time({
  open_dataset(
    sources = "data/seattle-library-checkouts.parquet", 
    format = "parquet"
  ) |>
    filter(CheckoutYear == 2021, MaterialType == "BOOK") |>
    group_by(CheckoutMonth) |>
    summarise(TotalCheckouts = sum(Checkouts)) |>
    arrange(desc(CheckoutMonth)) |>
    collect()
})
```

The analysis of the parquet data finishes in about one tenth the time it takes to perform the same analysis with the csv data.

## Multi-file data sets

When data sets become large it is often useful to split the data across many files rather than storing everything in a single large file.
When this structuring is done intelligently, the multi-file strategy can lead to significant improvements in performance.
We'll illustrate this using the Seattle library data and the arrow package.

### Writing multi-file data

When deciding to pursue a multi-file strategy, the first task is to decide how to partition your data set.
If you break the data set into too few files you won't gain the performance improvements you're hoping for, but if you break it into too many you won't gain much either.
If you split the data in a meaningful way, you'll get some performance advantages, but if you split the data randomly you probably won't see a lot of gains.
There are no hard and fast rules about how to partition a data set: the results will depend on your data, access patterns, and the systems that read the data.
As a rough guide, the arrow package documentation suggests the following heuristics:

-   Avoid files smaller than 20MB and larger than 2GB.
-   Avoid partitioning layouts that produce more than 10,000 distinct files.

According to these guidelines, the Seattle library data is a little too large to be stored as a single 4GB parquet file.
We should consider splitting it up.

As a general rule, a useful strategy for partitioning a large dataset is to create meaningful subsets based on one or more variable that you're likely to use to filter the data.
For the library data it seems likely that we would often want to analyze the data separately by year, so the `CheckoutYear` column seems like a good candidate.

Partitioning a data set can be done using the `write_dataset()` function from the arrow package, which respects the grouping created by the dplyr `group_by()` function.
To create a multi-file data set with one parquet file per year of library data we would do the following:

```{r write-dataset}
#| eval: !expr "!file.exists('data/seattle-library-checkouts/CheckoutYear=2021/part-0.parquet')"
seattle_lib |>
  group_by(CheckoutYear) |>
  write_dataset(
    path = "data/seattle-library-checkouts",
    format = "parquet"
  )
```

Let's take a look at the files:

```{r show-parquet-files}
seattle_lib_files <- list.files(
  path = "data/seattle-library-checkouts",
  recursive = TRUE
) 
seattle_lib_files
```

These filenames follow a "self-describing" convention used by the Apache Hive project ([hive.apache.org](https://hive.apache.org/)).
Hive-style partitions name folders using a "key=value" convention, and the data files in that folder contain the subset of the data for which the key has the relevant value.
Following this convention makes it easier to locate files relevant to a particular query.

### Performance advantages

Partitioning a large data set into many smaller, meaningful files can produce substantial improvements in performance.
To illustrate, we'll repeat the analysis of the 2021 Seattle library book checkouts by month using the partitioned data:

```{r dataset-performance-multiple-parquet}
#| cache: true
system.time({
  open_dataset(
    sources = "data/seattle-library-checkouts/", 
    format = "parquet"
  ) |>
    filter(CheckoutYear == 2021, MaterialType == "BOOK") |>
    group_by(CheckoutMonth) |>
    summarise(TotalCheckouts = sum(Checkouts)) |>
    arrange(desc(CheckoutMonth)) |>
    collect()
})
```

This is a 10-fold speed up compared to a single parquet file, and a 100-fold speed up compared to the original csv file.

## Other topics

I haven't quite worked out the best way to integrate these sections.

### Using duckdb with arrow

One useful feature of arrow is that it is interoperable with duckdb and dbplyr.
It is possible to open a multi-file data set using arrow and then pass it to duckdb using the `to_duckdb()` function.
Doing so means that your dplyr code will now be interpreted by dbplyr (rather than arrow) and executed by duckdb.

```{r use-duckdb}
open_dataset("data/seattle-library-checkouts/") |>
  to_duckdb() |>
  filter(CheckoutYear == 2021, MaterialType == "BOOK") |>
  group_by(CheckoutMonth) |>
  summarise(TotalCheckouts = sum(Checkouts)) |>
  arrange(desc(CheckoutMonth)) |>
  collect()
```

There is an analogous function `to_arrow()` that passes control back to arrow, meaning that it is possible to swap back and forth between the duckdb and arrow back ends within a single analysis pipeline.
This operation is extremely efficient (only the metadata need to be copied from one system to the other) can occasionally be valuable when one system implements functionality that the other does not.

### The arrow in-memory and on-disk format

The main focus in this chapter has been on the parquet format, but it is worth noting in passing that arrow defines its own binary data format that represents data on-disk in (almost) the exact format that it uses when representing rectangular data in-memory.
To unpack this a little, let's load the parquet file storing the 2021 Seattle library data:

```{r read-arrow-table}
"!file.exists('data/seattle-library-checkouts/CheckoutYear=2021/part-0.parquet')"
checkouts2021 <- read_parquet(
  file = "data/seattle-library-checkouts/CheckoutYear=2021/part-0.parquet",
  as_data_frame = FALSE
)
glimpse(checkouts2021)
```

Note that the first line of output tells us this is a Table object.
A Table is the Apache Arrow analog to a data frame: it is represented in-memory, and arrow computations are done using these objects.
For our purposes, the important thing to understand is that the arrow file format is designed to be an exact on-disk analog of an in-memory Table.
You can read and write these files using the `read_ipc_file()` and `write_ipc_file()` functions in the arrow package:

```{r write-arrow-file}
arrow_file <- tempfile(fileext = ".arrow")
write_ipc_file(checkouts2021, arrow_file)
```

Note that for future releases of arrow, these functions are likely to deprecated in favour of the names `read_arrow()` and `write_arrow()` respectively.
Also note that the arrow file format was previously known as the "feather" file format, and you are likely to see references to it by that name.

To learn more about Datasets, Tables, and other arrow data objects, see the documentation on the arrow package website: [arrow.apache.org/docs/r/](https://arrow.apache.org/docs/r/).

### Data in the cloud

Sometimes when working with large datasets you encounter the situation when data files are stored remotely on cloud storage services such as Amazon's Simple Storage Service (S3) or Google Cloud Storage (GCS).
Functions in arrow that read and write data, including `open_dataset()`, are able to read and write from remote data sources with the help of the `s3_bucket()` function (for Amazon S3) or the `gs_bucket()` function (for Google cloud).
An example is shown below:

```{r open-s3-dataset}
#| cache: true
diamonds <- s3_bucket("voltrondata-labs-datasets/diamonds")
diamonds |> 
  open_dataset() |>
  glimpse()
```
