# Data import {#data-import}

```{r, results = "asis", echo = FALSE}
status("restructuring")
```

<!--# TO DO: This chapter got moved here from the wrangle section, make sure it makes sense in this new location, doesn't assume anything that comes after it. -->

## Introduction

Working with data provided by R packages is a great way to learn the tools of data science, but at some point you want to stop learning and start working with your own data.
In this chapter, you'll learn how to read plain-text rectangular files into R.
Here, we'll only scratch the surface of data import, but many of the principles will translate to other forms of data.
We'll finish with a few pointers to packages that are useful for other types of data.

### Prerequisites

In this chapter, you'll learn how to load flat files in R with the **readr** package, which is part of the core tidyverse.

```{r setup, message = FALSE}
library(tidyverse)
```

## Getting started

Most of readr's functions are concerned with turning flat files into data frames:

-   `read_csv()` reads comma delimited files, `read_csv2()` reads semicolon separated files (common in countries where `,` is used as the decimal place), `read_tsv()` reads tab delimited files, and `read_delim()` reads in files with any delimiter.

-   `read_fwf()` reads fixed width files.
    You can specify fields either by their widths with `fwf_widths()` or their position with `fwf_positions()`.
    `read_table()` reads a common variation of fixed width files where columns are separated by white space.

-   `read_log()` reads Apache style log files.
    (But also check out [webreadr](https://github.com/Ironholds/webreadr) which is built on top of `read_log()` and provides many more helpful tools.)

These functions all have similar syntax: once you've mastered one, you can use the others with ease.
For the rest of this chapter we'll focus on `read_csv()`.
Not only are csv files one of the most common forms of data storage, but once you understand `read_csv()`, you can easily apply your knowledge to all the other functions in readr.

## Reading data from a file

Here is what a simple CSV file with a row for column names (also commonly referred to as the header row) and six rows of data looks like.

```{r fig.cap = "hello", echo = FALSE}
read_lines("data/students.csv") %>% cat(sep = "\n")
```

Note that the `,`s separate the columns.
Table \@ref(tab:students-table) shows a representation of the same data as a table.

```{r students-table, echo = FALSE}
read_csv("data/students.csv") %>%
  knitr::kable(caption = "Data from the students.csv file as a table.")
```

The first argument to `read_csv()` is the most important: it's the path to the file to read.

```{r, message = TRUE}
students <- read_csv("data/students.csv")
```

When you run `read_csv()` it prints out a column specification that gives the name and type of each column.
That's an important part of readr, which we'll come back to in Section \@ref(parsing-a-file) on parsing a file.

<!--# Update examples below to use students.csv file -->

You can also supply an inline csv file.
This is useful for experimenting with readr and for creating reproducible examples to share with others:

```{r}
read_csv("a,b,c
1,2,3
4,5,6")
```

In both cases `read_csv()` uses the first line of the data for the column names, which is a very common convention.
There are two cases where you might want to tweak this behaviour:

1.  Sometimes there are a few lines of metadata at the top of the file.
    You can use `skip = n` to skip the first `n` lines; or use `comment = "#"` to drop all lines that start with (e.g.) `#`.

    ```{r}
    read_csv("The first line of metadata
      The second line of metadata
      x,y,z
      1,2,3", skip = 2)

    read_csv("# A comment I want to skip
      x,y,z
      1,2,3", comment = "#")
    ```

2.  The data might not have column names.
    You can use `col_names = FALSE` to tell `read_csv()` not to treat the first row as headings, and instead label them sequentially from `X1` to `Xn`:

    ```{r}
    read_csv("1,2,3\n4,5,6", col_names = FALSE)
    ```

    (`"\n"` is a convenient shortcut for adding a new line. You'll learn more about it and other types of string escape in \[string basics\].)

    Alternatively you can pass `col_names` a character vector which will be used as the column names:

    ```{r}
    read_csv("1,2,3\n4,5,6", col_names = c("x", "y", "z"))
    ```

Another option that commonly needs tweaking is `na`: this specifies the value (or values) that are used to represent missing values in your file:

```{r}
read_csv("a,b,c\n1,2,.", na = ".")
```

This is all you need to know to read \~75% of CSV files that you'll encounter in practice.
You can also easily adapt what you've learned to read tab separated files with `read_tsv()` and fixed width files with `read_fwf()`.
To read in more challenging files, you'll need to learn more about how readr parses each column, turning them into R vectors.

### Compared to base R

If you've used R before, you might wonder why we're not using `read.csv()`.
There are a few good reasons to favour readr functions over the base equivalents:

-   They are typically much faster (\~10x) than their base equivalents.
    Long running jobs have a progress bar, so you can see what's happening.
    If you're looking for raw speed, try `data.table::fread()`.
    It doesn't fit quite so well into the tidyverse, but it can be quite a bit faster.

-   They produce tibbles, they don't convert character vectors to factors, use row names, or munge the column names.
    These are common sources of frustration with the base R functions.

-   They are more reproducible.
    Base R functions inherit some behaviour from your operating system and environment variables, so import code that works on your computer might not work on someone else's.

### Exercises

1.  What function would you use to read a file where fields were separated with\
    "\|"?

2.  Apart from `file`, `skip`, and `comment`, what other arguments do `read_csv()` and `read_tsv()` have in common?

3.  What are the most important arguments to `read_fwf()`?

4.  Sometimes strings in a CSV file contain commas.
    To prevent them from causing problems they need to be surrounded by a quoting character, like `"` or `'`. By default, `read_csv()` assumes that the quoting character will be `"`.
    What argument to `read_csv()` do you need to specify to read the following text into a data frame?

    ```{r, eval = FALSE}
    "x,y\n1,'a,b'"
    ```

5.  Identify what is wrong with each of the following inline CSV files.
    What happens when you run the code?

    ```{r, eval = FALSE}
    read_csv("a,b\n1,2,3\n4,5,6")
    read_csv("a,b,c\n1,2\n1,2,3,4")
    read_csv("a,b\n\"1")
    read_csv("a,b\n1,2\na,b")
    read_csv("a;b\n1;3")
    ```

## Writing to a file

readr also comes with two useful functions for writing data back to disk: `write_csv()` and `write_tsv()`.
Both functions increase the chances of the output file being read back in correctly by:

-   Always encoding strings in UTF-8.

-   Saving dates and date-times in ISO8601 format so they are easily parsed elsewhere.

If you want to export a csv file to Excel, use `write_excel_csv()` --- this writes a special character (a "byte order mark") at the start of the file which tells Excel that you're using the UTF-8 encoding.

The most important arguments are `x` (the data frame to save), and `path` (the location to save it).
You can also specify how missing values are written with `na`, and if you want to `append` to an existing file.

```{r, eval = FALSE}
write_csv(challenge, "challenge.csv")
```

Note that the type information is lost when you save to csv:

```{r, warning = FALSE}
challenge
write_csv(challenge, "challenge-2.csv")
read_csv("challenge-2.csv")
```

This makes CSVs a little unreliable for caching interim results---you need to recreate the column specification every time you load in.
There are two alternatives:

1.  `write_rds()` and `read_rds()` are uniform wrappers around the base functions `readRDS()` and `saveRDS()`.
    These store data in R's custom binary format called RDS:

    ```{r}
    write_rds(challenge, "challenge.rds")
    read_rds("challenge.rds")
    ```

2.  The feather package implements a fast binary file format that can be shared across programming languages:

    ```{r, eval = FALSE}
    library(feather)
    write_feather(challenge, "challenge.feather")
    read_feather("challenge.feather")
    #> # A tibble: 2,000 x 2
    #>       x      y
    #>   <dbl> <date>
    #> 1   404   <NA>
    #> 2  4172   <NA>
    #> 3  3004   <NA>
    #> 4   787   <NA>
    #> 5    37   <NA>
    #> 6  2332   <NA>
    #> # ... with 1,994 more rows
    ```

Feather tends to be faster than RDS and is usable outside of R.
RDS supports list-columns (which you'll learn about in Chapter \@ref(list-columns); feather currently does not.

```{r, include = FALSE}
file.remove("challenge-2.csv")
file.remove("challenge.rds")
```

## Other types of data

To get other types of data into R, we recommend starting with the tidyverse packages listed below.
They're certainly not perfect, but they are a good place to start.
For rectangular data:

-   **readxl** reads Excel files (both `.xls` and `.xlsx`).
    See Chapter \@ref(import-spreadsheets) for more on working with data stored in Excel spreadsheets.

-   **googlesheets4** reads Google Sheets.
    Also see Chapter \@ref(import-spreadsheets) for more on working with data stored in Google Sheets.

-   **DBI**, along with a database specific backend (e.g. **RMySQL**, **RSQLite**, **RPostgreSQL** etc) allows you to run SQL queries against a database and return a data frame.
    See Chapter \@ref(import-databases) for more on working with databases .

-   **haven** reads SPSS, Stata, and SAS files.

For hierarchical data: use **jsonlite** (by Jeroen Ooms) for json, and **xml2** for XML.
Jenny Bryan has some excellent worked examples at <https://jennybc.github.io/purrr-tutorial/>.

For other file types, try the [R data import/export manual](https://cran.r-project.org/doc/manuals/r-release/R-data.html) and the [**rio**](https://github.com/leeper/rio) package.
