# (PART) Explore {-}

# Introduction

```{r, include = FALSE}
library(ggplot2)
library(dplyr)
```

The human brain is not designed to work with raw data. Working memory can only attend to a few values at a time. This makes it difficult to discover patterns in raw data. For example, can you spot what is striking about the relationship between $X$ and $Y$ in the table below?

```{r data, echo=FALSE}
x <- rep(seq(0.2, 1.8, length = 5), 2) + runif(10, -0.15, 0.15)
X <- c(0.02, x, 1.94)
Y <- sqrt(1 - (X - 1)^2)
Y[1:6] <- -1 * Y[1:6]
Y <- Y - 1
order <- sample(1:10)
knitr::kable(round(data.frame(X = X[order], Y = Y[order]), 2))
```

While we may stumble over raw data, we can easily process visual information. Within the brain is a visual processing system that has been fine-tuned by thousands of years of evolution. This is why the quickest way to understand data is to visualize it. Once you plot your data, you can instantly see any relationship between values. Here, we see that the values above fall on a circle.

```{r echo=FALSE, dependson=data}
ggplot2::qplot(X, Y) + ggplot2::coord_fixed(ylim = c(-2.5, 2.5), xlim = c(-2.5, 2.5))
```

Visualization works because your brain processes visual information in a different (and much wider) channel than it processes symbolic information, like words and numbers. However, visualization is not the only way to comprehend data.

You can also comprehend data by transforming it. Because you can easily attend to a small number of values, you can use summary values, like averages, maximums, minimums, medians, to describe important aspects of the data.

Another way to summarize your data is to use a model, a function that describes the relationship between two or more variables (sets of values). For the reasons above, you can more easily attend to the parts of a model than you can to all the raw values in your data set.

The first problem in Data Science is a cognitive one: understanding your data. In this part of the book, you'll learn how to use R to accomplish this task.
