---
layout: default
title: Expressing yourself in code
---

# Expressing yourself in code

```{r, include = FALSE}
source("common.R")
knitr::opts_chunk$set(
  cache = TRUE,
  fig.path = "figures/functions/"
)
library(dplyr)
diamonds <- ggplot2::diamonds
```

Code is a tool of communication, not just to the computer, but to other people. This is important because every project you undertake is fundamentally collaborative. Even if you're not working with other people, you'll definitely be working with future-you. You want to write clear code so that future-you doesn't curse present-you when you look at a project again after several months have passed.

To me, improving your communication skills is a key part of mastering R as a programming language. Over time, you want your code to becomes more and more clear, and easier to write. In this chapter, you'll learn three important skills that help you to move in this direction: 

1.  We'll dive deep in to the __pipe__, `%>%`, talking more about how it works 
    and how it gives you a new tool for rewriting your code. You'll also learn 
    about when not to use the pipe!

1.  Repeating yourself in code is dangerous because it can easily lead to 
    errors and inconsistencies. We'll talk about how to write __functions__
    in order to remove duplication in your logic.
    
1.  Another important tool for removing duplication is the __for loop__ which
    allows you to repeat the same action again and again and again. You tend to 
    use for-loops less often in R than in other programming languages because R 
    is a functional programming language which means that you can extract out 
    common patterns of for loops and put them in a function. We'll come back to
    that idea in XYZ.

Removing duplication is an important part of expressing yourself clearly because it lets the reader (i.e. future-you!) focus on what's different between operations rather than what's the same. The goal is not just to write better funtions or to do things that you couldn't do before, but to code with more "ease". As you internalise the ideas in this chapter, you should find it easier to re-tackle problems that you've solved in the past with much effort.

Writing code is similar in many ways to writing prose. One parallel which I find particularly useful is that in both cases rewriting is key to clarity. The first expression of your ideas is unlikely to be particularly clear, and you may need to rewrite multiple times. After solving a data analysis challenge, it's often worth looking at your code and thinking about whether or not it's obvious what you've done. If you spend a little time rewriting your code while the ideas are fresh, you can save a lot of time later trying to recreate what your code did. But this doesn't mean you should rewrite every function: you need to balance what you need to achieve now with saving time in the long run.  (But the more you rewrite your functions the more likely you'll first attempt will be clear.)

## Piping

Pipes let you transform the way you call deeply nested functions. Using a pipe doesn't affect at all what the code does; behind the scenes it is run in exactly the same way. What the pipe does is change how the code is written and hence how it is read. It tends to transform to a more imperative form (do this, do that, do that other thing, ...)  so that it's easier to read.

### Piping alternatives

To explore how you can write the same code in many different ways, let's use code to tell a story about a little bunny named foo foo:

> Little bunny Foo Foo  
> Went hopping through the forest  
> Scooping up the field mice  
> And bopping them on the head  

We'll start by defining an object to represent litte bunny Foo Foo:

```{r, eval = FALSE}
foo_foo <- little_bunny()
```

And then we'll use a function for each key verb `hop()`, `scoop()`, and `bop()`. Using this object and these verbs, there are a number of ways we could retell the story in code:

* Save each intermediate step as a new object
* Rewrite the original object multiple times
* Compose functions
* Use the pipe

Below we work through each approach, showing you the code and talking about the advantages and disadvantages.

#### Intermediate steps

The simplest and most robust approach to sequencing multiple function calls is to save each intermediary as a new object:

```{r, eval = FALSE}
foo_foo_1 <- hop(foo_foo, through = forest)
foo_foo_2 <- scoop(foo_foo_1, up = field_mice)
foo_foo_3 <- bop(foo_foo_2, on = head)
```

The main downside of this form is that it forces you to name each intermediate element. If there are natural names, this form feels natural, and you should use it. But if you're giving then arbitrary unique names, like this example, I don't think it's that useful. Whenever I write code like this, I invariably write the wrong number somewhere and then spend 10 minutes scratching my head and trying to figure out what went wrong with my code.

You may worry that this form creates many intermediate copies of your data and takes up a lot of memory. First, in R, worrying about memory is not a useful way to spend your time: worry about it when it becomes a problem (i.e. you run out of memory), not before. Second, R isn't stupid: it will reuse the shared columns in a pipeline of data frame transformations. Let's take a look at an actual data manipulation pipeline where we add a new column to the `diamonds` dataset from ggplot2:

```{r}
diamonds2 <- mutate(diamonds, price_per_carat = price / carat)

library(pryr)
object_size(diamonds)
object_size(diamonds2)
object_size(diamonds, diamonds2)
```

`pryr::object_size()` gives the memory occupied by all of its arguments. The results seem counterintuitive at first:

* `diamonds` takes up 3.46 MB,
* `diamonds2` takes up 3.89 MB,
* `diamonds` and `diamonds2` together take up 3.89 MB!

How can that work? Well, `diamonds2` has 10 columns in common with `diamonds`: there's no need to duplicate all that data so both data frames share the vectors. R will only create a copy of a vector if you modify it. Modifying a single value will mean that the data frames can no longer share as much memory. The individual sizes will be unchange, but the collective size will increase:

```{r}
diamonds$carat[1] <- NA
object_size(diamonds)
object_size(diamonds2)
object_size(diamonds, diamonds2)
```

(Note that we use `pryr::object_size()` here, not the built-in `object.size()`, because it doesn't have quite enough smarts.)

#### Overwrite the original

One way to eliminate the intermediate objects is to just overwrite the same object again and again:

```{r, eval = FALSE}
foo_foo <- hop(foo_foo, through = forest)
foo_foo <- scoop(foo_foo, up = field_mice)
foo_foo <- bop(foo_foo, on = head)
```

This is less typing (and less thinking), so you're less likely to make mistakes. However, there are two problems:

1.  It will make debugging painful: if you make a mistake you'll need to start 
    again from scratch. 
    
1.  The reptition of the object being transformed (we've written `foo_foo` six 
    times!) obscures what's changing on each line. 
    
#### Function composition

Another approach is to abandon assignment altogether and just string the function calls together:

```{r, eval = FALSE}
bop(
  scoop(
    hop(foo_foo, through = forest),
    up = field_mice
  ), 
  on = head
)
```

Here the disadvantage is that you have to read from inside-out, from right-to-left, and that the arguments end up spread far apart (sometimes called the 
[dagwood sandwhich](https://en.wikipedia.org/wiki/Dagwood_sandwich) problem).

#### Use the pipe 

Finally, we can use the pipe:

```{r, eval = FALSE}
foo_foo %>%
  hop(through = forest) %>%
  scoop(up = field_mouse) %>%
  bop(on = head)
```

This is my favourite form. The downside is that you need to understand what the pipe does, but once you've mastered that idea task, you can read this series of function compositions like it's a set of imperative actions. Foo foo, hops, then scoops, then bops.

Behind the scenes magrittr converts this to:

```{r, eval = FALSE}
. <- hop(foo_foo, through = forest)
. <- scoop(., up = field_mice)
bop(., on = head)
```

It's useful to know this because if an error is thrown in the middle of the pipe, you'll need to be able to interpret the `traceback()`.

### Other tools from magrittr

The pipe is provided by the magrittr package, by Stefan Milton Bache. Most of packages you work in this book automatically provide `%>%` for you. You might want to load magrittr yourself if you're using another package, or you want to access some of the other pipe variants that magrittr provides.

```{r}
library(magrittr)
```

*   When working with more complex pipes, it's some times useful to call a 
    function for its side-effects. Maybe you want to print out the current 
    object, or plot it, or save it to disk. Many times, such functions don't 
    return anything, effectively terminating the pipe.
    
    To work around this problem, you can use the "tee" pipe. `%T>%` works like 
    `%>%` except instead it returns the LHS instead of the RHS. It's called 
    "tee" because it's like a literal T-shaped pipe.

    ```{r}
    rnorm(100) %>%
      matrix(ncol = 2) %>%
      plot() %>%
      str()
    
    rnorm(100) %>%
      matrix(ncol = 2) %T>%
      plot() %>%
      str()
    ```

*   If you're working with functions that don't have a dataframe based API  
    (i.e. you pass them individual vectors, not a data frame and expressions 
    to be evaluated in the context of that data frame), you might find `%$%` 
    useful. It "explodes" out the variables in a data frame so that you can 
    refer to them explicitly. This is useful when working with many functions 
    in base R:
    
    ```{r}
    mtcars %$%
      cor(disp, mpg)
    ```

*   For assignment. magrittr provides the `%<>%` operator which allows you to
    replace code like:
  
    ```R
    mtcars <- mtcars %>% transform(cyl = cyl * 2)
    ```
    
    with
     
    ```R
    mtcars %<>% transform(cyl = cyl * 2)
    ```
    
    I'm not a fan of this operator because I think assignment is such a 
    special operation that it should always be clear when it's occuring.
    In my opinion, a little bit of duplication (i.e. repeating the 
    name of the object twice), is fine in return for making assignment
    more explicit.

### When not to use the pipe

The pipe is a powerful tool, but it's not the only tool at your disposal, and it doesn't solve every problem! Pipes are most useful for rewriting a fairly short linear sequence of operations. I think you should reach for another tool when:

* Your pipes get longer than five or six lines. In that case, create 
  intermediate objects with meaningful names. That will make debugging easier,
  because you can more easily check the intermediate results. It also helps
  when reading the code, because the variable names can help describe the
  intent of the code.
  
* You have multiple inputs or outputs. If there is not one primary object
  being transformed, write code the regular ways.

* You are start to think about a directed graph with a complex
  dependency structure. Pipes are fundamentally linear and expressing 
  complex relationships with them typically does not yield clear code.

### Pipes in production

When you run a pipe interactively, it's easy to see if something goes wrong. When you start writing pipes that are used in production, i.e. they're run automatically and a human doesn't immediately look at the output it's a really good idea to include some assertions that verify the data  looks like expect. One great way to do this is the ensurer package, writen by Stefan Milton Bache (the author of magrittr). 
  
<http://www.r-statistics.com/2014/11/the-ensurer-package-validation-inside-pipes/>

## Functions

Whenever you've copied and pasted code more than twice, you need to take a look at it and see if you can extract out the common components and make a function. For example, take a look at this code. What does it do?

```{r}
df <- data.frame(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

df$a <- (df$a - min(df$a, na.rm = TRUE)) / 
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
df$b <- (df$b - min(df$b, na.rm = TRUE)) / 
  (max(df$a, na.rm = TRUE) - min(df$b, na.rm = TRUE))
df$c <- (df$c - min(df$c, na.rm = TRUE)) / 
  (max(df$c, na.rm = TRUE) - min(df$c, na.rm = TRUE))
df$d <- (df$d - min(df$d, na.rm = TRUE)) / 
  (max(df$d, na.rm = TRUE) - min(df$d, na.rm = TRUE))
```

You might be able to puzzle out that this rescales each column to 0--1. But did you spot the mistake? I made an error when updating the code for `df$b`, and I forgot to change an `a` to a `b`. Extracting repeated code out into a function is a good idea because it helps make your code more understandable (because you can name the operation), and it prevents you from making this sort of copy-and-paste error.

To write a function you need to first analyse the operation. How many inputs does it have?

```{r, eval = FALSE}
(df$a - min(df$a, na.rm = TRUE)) /
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
```

It's often a good idea to rewrite the code using some temporary values. Here this function only takes one input, so I'll call it `x`:

```{r}
x <- 1:10
(x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
```

There is some duplication in this code: I'm computing the `min()` and `max()` multiple times, and I could instead do that in one step:

```{r}
rng <- range(x, na.rm = TRUE)
(x - rng[1]) / (rng[2] - rng[1])
```

Now that I've simplified the code, and made sure it works, I can turn it into a function:

```{r}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
rescale01(c(0, 5, 10))
```

Always make sure your code works on a simple test case before creating the function!

Note the process that I followed here: constructing the `function` is the last thing I did. It's much easier to start with code that works on a sample input and then turn it into a function rather than the other way around. You're more likely to get to your final destination if you take small steps and check your work after each step.

Now we can use that to simplify our original example:

```{r}
df$a <- rescale01(df$a)
df$b <- rescale01(df$b)
df$c <- rescale01(df$c)
df$d <- rescale01(df$d)
```

This makes it more clear what we're doing, and avoids one class of copy-and-paste errors. However, we still have quite a bit of duplication: we're still doing the same thing to multiple columns. We'll learn how to handle that in the for loop section. But first, lets talk a bit more about functions.

### Practice

Practice turning the following code snippets into functions. Think about how you can re-write them to be as clear an expressive as possible.

### Function components

There are three attributes that define what a function does:

1.  The __arguments__ of a function are its inputs. 

1.  The __body__ of a function is the code that it runs each time.

1.  The function __environment__ controls how it looks up values from names
    (i.e. how it goes from the name `x`, to its value, `10`).

#### Arguments

You can choose to supply default values to your arguments for common options. This is useful so that you don't need to repeat yourself all the time.

```{r}
foo <- function(x = 1, y = TRUE, z = 10:1) {
  
}
```

Default values can depend on other arguments but don't over use this technique as it's possible to create code that is very difficult to understand:

```{r}
bar <- function(x = y + 1, y = x + 1) {
  x * y
}
```

On other aspect of arguments you'll commonly see is `...`. This captures any other arguments not otherwise matched. It's useful because you can then send those `...` on to another argument. This is a useful catch all if your function primarily wraps another function.  For example, you might have written your own wrapper designed to add linear model lines to a ggplot:

```{r}
geom_lm <- function(formula = y ~ x, colour = alpha("steelblue", 0.5), 
                    size = 2, ...)  {
  geom_smooth(formula = formula, se = FALSE, method = "lm", colour = colour,
    size = size, ...)
}
```

This allows you to use any other arguments of `geom_smooth()`, even thoses that aren't explicitly listed in your wrapper (and even arguments that don't exist yet in the version of ggplot2 that you're using).

#### Body

The body of the function does the actual work. The return value of a function is the last thing that it does. 

You can use an explicit `return()` statement, but this is not needed, and is best avoided except when you want to return early.

#### Environment

The environment of a function control where values are looked up from. Take this function for example:

```{r}
f <- function(x) {
  x + y
} 
```

In many programming languages, this would be an error, because `y` is not defined inside the function. However, in R this is valid code. Since `y` is not defined inside the function, R will look in the environment where the function was defined:

```{r}
y <- 100
f(10)

y <- 1000
f(10)
```

You should avoid functions that work like this because it makes it harder to predict what your function will return. 

This behaviour seems like a recipe for bugs, but by and large it doesn't cause too many, especially as you become a more experienced R programmer. The advantage of this behaviour is from a language stand point it allows R to be very consistent. Every name is looked up using the same set of rules. For `f()` that includes the behaviour of two things that you might not expect: `{` and `+`.

This consistent set of rules allows for a number of powerful tool that are unfortunately beyond the scope of this book, but you can read about in "Advanced R".

#### Exercises

1.  What happens if you try to override the method in `geom_lm()` created
    above? Why?

### Making functions with magrittr

One cool feature of the pipe is that it's also very easy to create functions with it.

### Non-standard evaluation

One challenge with writing functions is that many of the functions you've used in this book use non-standard evaluation to minimise typing. This makes these functions great for interactive use, but it does make it more challenging to program with them, because you need to use more advanced techniques.

Unfortunately these techniques are beyond the scope of this book, but you can learn the techniques with online resources:

* Programming with ggplot2 (an excerpt from the ggplot2 book):
  http://rpubs.com/hadley/97970
  
* Programming with dplyr: still hasn't been written.

### Exercises

1.  Follow <http://nicercode.github.io/intro/writing-functions.html> to 
    write your own functions to compute the variance and skew of a vector.

1.  Read the [complete lyrics](https://en.wikipedia.org/wiki/Little_Bunny_Foo_Foo) 
    to "Little Bunny Foo". There's a lot of duplication in this song. 
    Extend the initial piping example to recreate the complete song, using 
    functions to reduce duplication.

## For loops

Before we tackle the problem of rescaling each column, lets start with a simpler case. Imagine we want to summarise each column with its median. One way to do that is to use a for loop. Every for loop has three main components:

```{r}
results <- vector("numeric", ncol(df))
for (i in seq_along(df)) {
  results[[i]] <- median(df[[i]])
}
results
```

There are three parts to a for loop:

1.  The __results__: `results <- vector("integer", length(x))`. 
    This creates an integer vector the same length as the input. It's important
    to enough space for all the results up front, otherwise you have to grow the 
    results vector at each iteration, which is very slow for large loops.

1.  The __sequence__: `i in seq_along(df)`. This determines what to loop over:
    each run of the for loop will assign `i` to a different value from 
    `seq_along(df)`, shorthand for `1:length(df)`. It's useful to think of `i`
    as a pronoun.
    
1.  The __body__: `results[i] <- median(df[[i]])`. This code is run repeatedly, 
    each time with a different value in `i`. The first iteration will run 
    `results[1] <- median(df[[2]])`, the second `results[2] <- median(df[[2]])`, 
    and so on.

This loop used a function you might not be familiar with: `seq_along()`. This is a safe version of the more familiar `1:length(l)`. There's one important difference in behaviour. If you have a zero-length vector, `seq_along()` does the right thing:

```{r}
y <- numeric(0)
seq_along(y)
1:length(y)
```

Lets go back to our original motivation:

```{r}
df$a <- rescale01(df$a)
df$b <- rescale01(df$b)
df$c <- rescale01(df$c)
df$d <- rescale01(df$d)
```

In this case the output is already present: we're modifying an existing object. 

Need to think about a data frame as a list of column (we'll make this definition precise later on). The length of a data frame is the number of columns. To extract a single column, you use `[[`.

That makes our for loop quite simple:

```{r}
for (i in seq_along(df)) {
  df[[i]] <- rescale01(df[[i]])
}
```

For loops are not as important in R as they are in other languages as rather than writing your own for loops, you'll typically use prewritten functions that wrap up common for-loop patterns. You'll learn about those in the next chapter.  These functions are important because they wrap up the book-keeping code related to the for loop, focussing purely on what's happening. For example the two for-loops we wrote above can be rewritten as:

```{r}
library(purrr)

map_dbl(df, median)
df[] <- map(df, rescale01)
```

The focus is now on the function doing the modification, rather than the apparatus of the for-loop.

### Looping patterns

There are three basic ways to loop over a vector:

1.  Loop over the elements: `for (x in xs)`. Most useful for side-effects,
    but it's difficult to save the output efficiently.

1.  Loop over the numeric indices: `for (i in seq_along(xs))`. Most common
    form if you want to know the element (`xs[[i]]`) and it's position.

1.  Loop over the names: `for (nm in names(xs))`. Gives you both the name
    and the position. This is useful if you want to use the name in a
    plot title or a file name.

The most general form uses `seq_along(xs)`, because from the position you can access both the name and the value:

```{r, eval = FALSE}
for (i in seq_along(x)) {
  name <- names(x)[[i]]
  value <- x[[i]]
}
```

### Exercises    

1.  It's common to see for loops that don't preallocate the output and instead
    increase the length of a vector at each step:
    
    ```{r}
    results <- vector("integer", 0)
    for (i in seq_along(x)) {
      results <- c(results, lengths(x[[i]]))
    }
    results
    ```
    
    How does this affect performance? 

## Learning more

As you become a better R programmer, you'll learn more techniques for reducing various types of duplication. This allows you to do more with less, and allows you to express yourself more clearly by taking advantage of powerful programming constructs.

To learn more you need to study R as a programming language, not just an interactive environment for data science. We have written two books that will help you do so:

* [Hands on programming with R](http://shop.oreilly.com/product/0636920028574.do),
  by Garrett Grolemund. This is an introduction to R as a programming language 
  and is a great place to start if R is your first programming language.
  
* [Advanced R](http://adv-r.had.co.nz) by Hadley Wickham. This dives into the
  details of R the programming language. This is a great place to start if
  you've programmed in other languages and you want to learn what makes R 
  special, different, and particularly well suited to data analysis.
