# Regular expressions

```{r, results = "asis", echo = FALSE}
status("restructuring")
```

## Introduction

You learned the basics of regular expressions in Chapter \@ref(strings), but because regular expressions are a miniature language it's worth spending some extra time on the details.

The chapter starts by expanding your knowledge of patterns, to cover six important new topics (escaping, anchoring, character classes, shorthand classes, quantifiers, and alternation).
Here we'll focus mostly on the language itself, not the functions that use it.
That means we'll mostly work with toy character vectors, showing the results with `str_view()` and `str_view_all()`.
You'll need to take what you learn here and apply it to data frames with tidyr functions or by combining dplyr and stringr functions.

Next we'll talk about the important concepts of "grouping" and "capturing" which give you new ways to extract variables out of strings using `tidyr::separate_group()`.
Grouping also allows you to use back references which allow you do things like match repeated patterns.

We'll finish by discussing the various "flags" that allow you to tweak the operation of regular expressions and then cover a details about how regular expressions work that , and then discuss some useful strategies .

### Prerequisites

This chapter will use regular expressions as provided by the **stringr** package.

```{r setup, message = FALSE}
library(tidyverse)
```

It's worth noting that the regular expressions used by stringr are very slightly different to those of base R.
That's because stringr is built on top of the [stringi package](https://stringi.gagolewski.com), which is in turn built on top of the [ICU engine](https://unicode-org.github.io/icu/userguide/strings/regexp.html), whereas base R functions (like `gsub()` and `grepl()`) use either the [TRE engine](https://github.com/laurikari/tre) or the [PCRE engine](https://www.pcre.org).
Fortunately, the basics of regular expressions are so well established that you're unlikely to encounter any differences when working with the patterns you'll learn in this book.
You only need to be aware of the difference when you start to rely on advanced features like complex Unicode character ranges or special features that use the `(?…)` syntax.
You can learn more about these advanced features in `vignette("regular-expressions", package = "stringr")`.

Another useful reference is [https://www.regular-expressions.info/](https://www.regular-expressions.info/tutorial.html).
It's not R specific, but it includes a lot more information about how regular expressions actually work.

### Exercises

1.  Explain why each of these strings don't match a `\`: `"\"`, `"\\"`, `"\\\"`.

2.  How would you match the sequence `"'\`?

3.  What patterns will the regular expression `\..\..\..` match?
    How would you represent it as a string?

## Pattern language

You learned the very basics of the regular expression pattern language in Chapter \@ref(strings), and now its time to dig into more of the details.
First, we'll start with **escaping**, which allows you to match characters that the pattern language otherwise treats specially.
Next you'll learn about **anchors**, which allow you to match the start or end of the string.
Then you'll learn about **character classes** and their shortcuts, which allow you to match any character from a set.
We'll finish up with **quantifiers**, which control how many times a pattern can match, and **alternation**, which allows you to match either *this* or *that.*

The terms I use here are the technical names for each component.
They're not always the most evocative of their purpose, but it's very helpful to know the correct terms if you later want to Google for more details.

### Escaping {#regexp-escaping}

In Chapter \@ref(strings), you'll learned how to match a literal `.` by using `fixed(".")`.
What if you want to match a literal `.` as part of a regular expression?
You'll need to use an escape, which tells the regular expression you want it to match exactly, not use its special behavior.
Like strings, regexps use the backslash, `\`, to escape special behavior.
So to match a `.`, you need the regexp `\.`.
Unfortunately this creates a problem.
We use strings to represent regular expressions, and `\` is also used as an escape symbol in strings.
So to create the regular expression `\.` we need the string `"\\."`.

```{r}
# To create the regular expression \., we need to use \\.
dot <- "\\."

# But the expression itself only contains one:
str_view(dot)

# And this tells R to look for an explicit .
str_view(c("abc", "a.c", "bef"), "a\\.c")
```

In this book, I'll write regular expression as `\.` and strings that represent the regular expression as `"\\."`.

If `\` is used as an escape character in regular expressions, how do you match a literal `\`?
Well you need to escape it, creating the regular expression `\\`.
To create that regular expression, you need to use a string, which also needs to escape `\`.
That means to match a literal `\` you need to write `"\\\\"` --- you need four backslashes to match one!

```{r}
x <- "a\\b"
str_view(x)
str_view(x, "\\\\")
```

Alternatively, you might find it easier to use the raw strings you learned about in Section \@ref(raw-strings)).
That allows you to avoid one layer of escaping:

```{r}
str_view(x, r"(\\)")
```

The full set of characters with special meanings that need to be escaped is `.^$\|*+?{}[]()`.
In general, look at punctuation character with suspicion; if your regular expression isn't matching what you think it should, check if you've used any of these characters.

### Anchors

By default, regular expressions will match any part of a string.
If you want to match at the start of end you need to **anchor** the regular expression using `^` or `$`.

-   `^` to match the start of the string.
-   `$` to match the end of the string.

```{r}
x <- c("apple", "banana", "pear")
str_view(x, "a")  # match "a" anywhere
str_view(x, "^a") # match "a" at start
str_view(x, "a$") # match "a" at end
```

To remember which is which, try this mnemonic which I learned from [Evan Misshula](https://twitter.com/emisshula/status/323863393167613953): if you begin with power (`^`), you end up with money (`$`).
It's tempting to put `$` at the start, because that's how we write sums of money, but it's not what regular expressions want.

To force a regular expression to only match the full string, anchor it with both `^` and `$`:

```{r}
x <- c("apple pie", "apple", "apple cake")
str_view(x, "apple")
str_view(x, "^apple$")
```

You can also match the boundary between words with `\b`.
I don't often use this in my R code, but I'll sometimes use it when I'm doing a search in RStudio.
It's use to find the name of a function that's a component of other functions.
For example, I'll search for `\bsum\b` to avoid matching `summarise`, `summary`, `rowsum` and so on:

```{r}
x <- c("summary(x)", "summarise(df)", "rowsum(x)", "sum(x)")
str_view(x, "sum")
str_view(x, "\\bsum\\b")
```

### Character classes

You can also create your own collections of characters using `[]`:

-   `[abc]`: matches a, b, or c.
-   `[a-z]`: matches every character between a and z. `[0-9]` matches any number.
-   `[^abc]`: matches anything except a, b, or c.
-   `[\^\-]`: matches `^` or `-`.

Remember that these are case sensitive.
If you want to match any Latin letter or number, you could write `[a-zA-Z0-9]`.

A character class containing a single character can be a nice alternative to escapes when you want to include a single special character (i.e. `$` `.` `|` `?` `*` `+` `(` `)` `[` `{` `}`, but not `]` `\` `^`).
This can be more readable because there are fewer slashes, but it does require a deeper understanding of regular expressions.

```{r}
# Look for a literal character that normally has special meaning in a regex
str_view(c("abc", "a.c", "a*c", "a c"), "a[.]c")
str_view(c("abc", "a.c", "a*c", "a c"), ".[*]c")
str_view(c("abc", "a.c", "a*c", "a c"), "a[ ]")
```

### Shorthand character classes

There are a few character classes that are used so commonly that they get their own shortcut.
You've already seen `.`, which matches any character apart from a newline.
There are three other useful pairs:

-   `\d`: matches any digit; `\D` matches anything that isn't a digit.
-   `\s`: matches any whitespace (e.g. space, tab, newline); `\S` matches anything that isn't whitespace.
-   `\w` matches any "word" character, i.e. letters and numbers; `\W`, matches any non-word character.

Remember, to create a regular expression containing `\d` or `\s`, you'll need to escape the `\` for the string, so you'll type `"\\d"` or `"\\s"`.
The following code demonstrates the different matches with a selection of letters, numbers, and punctuation characters.

```{r}
str_view_all("abcd12345!@#%. ", "\\d+")
str_view_all("abcd12345!@#%. ", "\\D+")
str_view_all("abcd12345!@#%. ", "\\w+")
str_view_all("abcd12345!@#%. ", "\\W+")
str_view_all("abcd12345!@#%. ", "\\s+")
str_view_all("abcd12345!@#%. ", "\\S+")
```

### Quantifiers

The next step up in power involves controlling how many times a pattern matches, the so called **quantifiers**.
We discussed `?` (0 or 1 matches), `+` (1 or more matches), and `*` (0 or more matches) in the last chapter.
Note that the precedence of these operators is high, so you can write: `colou?r` to match either American or British spellings.
That means most uses will need parentheses, like `bana(na)+`.

You can also specify the number of matches precisely:

-   `{n}`: exactly n
-   `{n,}`: n or more
-   `{n,m}`: between n and m

```{r}
x <- "1888 is the longest year in Roman numerals: MDCCCLXXXVIII"
str_view(x, "C{2}")
str_view(x, "C{2,}")
str_view(x, "C{1,3}")
str_view(x, "C{2,3}")
```

By default these matches are **greedy**: they will match the longest string possible.
You can make them **lazy**, matching the shortest string possible by putting a `?` after them.
This is an advanced feature of regular expressions, but it's useful to know that it exists:

```{r}
str_view(x, 'C{2,3}?')
str_view(x, 'C+[LX]+')
str_view(x, 'C+[LX]+?')
```

### Alternation

You can use **alternation** to pick between one or more alternative patterns.
This is a more general form of character classes that's not limited to match single characters.
I recommend always pairing `|` with parentheses, to make it very clear what the alternatives are.
Here are a few examples:

-   Match apple, pear, or banana: `"(apple)|(pear)|(banana)"`
-   Match 3 letters or two digits: `"(\\w{3})|(\\d{3})"`

We'll come back to parentheses very shortly in more detail.

For example, `abc|def` will match either `"abcef"`, or `"abdef"`.
Note that the precedence for `|` is low, so you'll often need to use it with parentheses: `(abc)|(def)` will match either `"abc"`, or `"def"`.
`abc|xyz` matches `abc` or `xyz` not `abcyz` or `abxyz`.
Like with mathematical expressions, if precedence ever gets confusing, use parentheses to make it clear what you want:

```{r}
str_view(c("grey", "gray"), "gr(e|a)y")
```

### Exercises

1.  How would you match the literal string `"$^$"`?

2.  Given the corpus of common words in `stringr::words`, create regular expressions that find all words that:

    a.  Start with "y".
    b.  Don't start with "y".
    c.  End with "x".
    d.  Are exactly three letters long. (Don't cheat by using `str_length()`!)
    e.  Have seven letters or more.

    Since `words` is long, you might want to use the `match` argument to `str_view()` to show only the matching or non-matching words.

3.  What strings will `$a` match?

4.  Create regular expressions to find all words that:

    a.  Start with a vowel.
    b.  That only contain consonants. (Hint: thinking about matching "not"-vowels.)
    c.  End with `ed`, but not with `eed`.
    d.  End with `ing` or `ise`.

5.  Empirically verify the rule "i before e except after c".

6.  Is "q" always followed by a "u"?

7.  Write a regular expression that matches a `word` if it's probably written in British English, not American English.

8.  Create a regular expression that will match telephone numbers as commonly written in your country.

9.  Describe the equivalents of `?`, `+`, `*` in `{m,n}` form.

10. Describe in words what these regular expressions match: (read carefully to see if I'm using a regular expression or a string that defines a regular expression.)

    a.  `^.*$`
    b.  `"\\{.+\\}"`
    c.  `\d{4}-\d{2}-\d{2}`
    d.  `"\\\\{4}"`

11. Create regular expressions to find all words that:

    a.  Start with three consonants.
    b.  Have three or more vowels in a row.
    c.  Have two or more vowel-consonant pairs in a row.

12. Solve the beginner regexp crosswords at <https://regexcrossword.com/challenges/beginner>.

## Grouping and capturing

Earlier, you learned about parentheses as a way to disambiguate complex expressions.
Parentheses also create a numbered capturing group (number 1, 2 etc.).
A capturing group stores the part of the string matched by the part of the regular expression inside the parentheses.
You can refer to the same text as previously matched by a capturing group with **backreferences**, like `\1`, `\2` etc.

For example, the following regular expression finds all fruits that have a repeated pair of letters.

```{r}
str_view(fruit, "(..)\\1", match = TRUE)
```

### Replacement

You can also use backreferences when replacing.
The following code will switch the order of the second and third words:

```{r}
sentences %>% 
  str_replace("(\\w+) (\\w+) (\\w+)", "\\1 \\3 \\2") %>% 
  head(5)
```

Names that start and end with the same letter.
Implement with `str_sub()` instead.

### str_match()

```{r}
sentences %>% 
  str_view("the (\\w+) (\\w+)", match = TRUE) %>% 
  head()
```

### Non-capturing groups

Occasionally, you'll want to use parentheses without creating matching groups.
You can create a non-capturing group with `(?:)`.
Typically, however, you'll find it easier to just ignore that result in the output of `str_match()`.

```{r}
x <- c("a gray cat", "a grey dog")
str_match(x, "(gr(e|a)y)")
str_match(x, "(gr(?:e|a)y)")
```

### Exercises

1.  Describe, in words, what these expressions will match:

    a.  `(.)\1\1`
    b.  `"(.)(.)\\2\\1"`
    c.  `(..)\1`
    d.  `"(.).\\1.\\1"`
    e.  `"(.)(.)(.).*\\3\\2\\1"`

2.  Construct regular expressions to match words that:

    a.  Start and end with the same character.
    b.  Contain a repeated pair of letters (e.g. "church" contains "ch" repeated twice.)
    c.  Contain one letter repeated in at least three places (e.g. "eleven" contains three "e"s.)

## Flags

The are a number of settings, called **flags**, that you can use to control some of the details of the pattern language.
In stringr, you can supply these by instead of passing a simple string as a pattern, by passing the object created by `regex()`:

```{r, eval = FALSE}
# The regular call:
str_view(fruit, "nana")
# Is shorthand for
str_view(fruit, regex("nana"))
```

This is useful because it allows you to pass additional arguments to control the details of the match the most useful is probably `ignore_case = TRUE` because it allows characters to match either their uppercase or lowercase forms:

```{r}
bananas <- c("banana", "Banana", "BANANA")
str_view(bananas, "banana")
str_view(bananas, regex("banana", ignore_case = TRUE))
```

If you're doing a lot of work with multiline strings (i.e. strings that contain `\n`), `multiline` and `dotall` can also be useful.
`dotall = TRUE` allows `.` to match everything, including `\n`:

```{r}
x <- "Line 1\nLine 2\nLine 3"
str_view_all(x, ".L")
str_view_all(x, regex(".L", dotall = TRUE))
```

And `multiline = TRUE` allows `^` and `$` to match the start and end of each line rather than the start and end of the complete string:

```{r}
x <- "Line 1\nLine 2\nLine 3"
str_view_all(x, "^Line")
str_view_all(x, regex("^Line", multiline = TRUE))
```

If you're writing a complicated regular expression and you're worried you might not understand it in the future, `comments = TRUE` can be super useful.
It allows you to use comments and white space to make complex regular expressions more understandable.
Spaces and new lines are ignored, as is everything after `#`.
(Note that I'm using a raw string here to minimise the number of escapes needed)

```{r}
phone <- regex(r"(
  \(?     # optional opening parens
  (\d{3}) # area code
  [) -]?  # optional closing parens, space, or dash
  (\d{3}) # another three numbers
  [ -]?   # optional space or dash
  (\d{3}) # three more numbers
  )", comments = TRUE)

str_match("514-791-8141", phone)
```

If you're using comments and want to match a space, newline, or `#`, you'll need to escape it:

```{r}
str_view("x x #", regex("x #", comments = TRUE))
str_view("x x #", regex(r"(x\ \#)", comments = TRUE))
```

## Some details

### Overlapping

Matches never overlap, and the regular expression engine only starts looking for a new match after the end of the last match.
For example, in `"abababa"`, how many times will the pattern `"aba"` match?
Regular expressions say two, not three:

```{r}
str_count("abababa", "aba")
str_view_all("abababa", "aba")
```

### Zero width matches

It's possible for a regular expression to match no character, i.e. the space between too characters.
This typically happens when you use a quantifier that allows zero matches:

```{r}
str_view_all("abcdef", "c?")
```

But anchors also create zero-width matches:

```{r}
str_view_all("this is a sentence", "\\b")
str_view_all("this is a sentence", "^")
```

### Greediness

Regular expressions always attempt to match the longest possible string.
