---
title: "Review Dashboard"
output: 
  flexdashboard::flex_dashboard:
    orientation: columns
---

```{r include = FALSE}
library(viridis)
library(ggplot2)
library(marmap)
```

# Intro {.sidebar}

This dashboard covers several topics:

* The marmap package
* The viridis package
* Miscellaneous material

# Marmap Package

## Column 1

### Florida

```{r echo = FALSE}
data(florida)
autoplot(florida)
```

The [marmap](https://cran.r-project.org/web/packages/marmap/index.html) package provides tools and data for visualizing the ocean floor. Here is an example contour plot of marmap's `florida` dataset.

## Column 2

### Hawaii

```{r echo = FALSE}
data(hawaii)
autoplot(hawaii)
```

### Alaska

```{r echo = FALSE}
data(aleutians)
autoplot(aleutians)
```


# Viridis Package

## Column 1

### Viridis colors

```{r fig.cap="Maunga Whao, Auckland, NZ"}
image(volcano, col = viridis(200))
```

### Magma colors

```{r fig.cap="Maunga Whao, Auckland, NZ"}
image(volcano, col = viridis(200, option = "A"))
```

## Column 2

### Inferno colors

```{r fig.cap="Maunga Whao, Auckland, NZ"}
image(volcano, col = viridis(200, option = "B"))
```

### Plasma colors

```{r fig.cap="Maunga Whao, Auckland, NZ"}
image(volcano, col = viridis(200, option = "C"))
```

# Miscellaneous

## Column 1 {data-width=300}

### Bash support

```{bash}
ls *.Rmd
```

***

This chunk executes bash code.

### Python support

```{python}
x = 'hello, python world!'
print(x.split(' '))
```

***

This chunk executes python code.

## Column 2 {data-width=700}

### Tables with Kable 

```{r echo = FALSE, results = 'asis'}
library(knitr)
kable(mtcars[1:5, ], caption = "A knitr kable.")
```

***

It is very easy to make tables with knitr's `kable` function.
